Assessing linkages between watershed nutrient loading and estuary water quality in Lavaca Bay, Texas

### 2023-02-11 Schramm-Manuscript-2023/Schramm-Manuscript-2023.tex
spell checks

% Abstract (Do not insert blank lines, i.e. \\)
\abstract{A single paragraph of about 200 words maximum. For research
articles, abstracts should give a pertinent overview of the work. We
strongly encourage authors to use the following style of structured
abstracts, but without headings: 1) Background: Place the question
addressed in a broad context and highlight the purpose of the study; 2)
Methods: Describe briefly the main methods or treatments applied; 3)
Results: Summarize the article's main findings; and 4) Conclusion:
Indicate the main conclusions or interpretations. The abstract should be
an objective representation of the article, it must not contain results
which are not presented and substantiated in the main text and should
not exaggerate the main conclusions.}

@@

Like many coastal areas globally, coastal watersheds along the Texas
Gulf coast are facing pressures from increasing population, increases in
point source and non-point source pollution and alterations to
freshwater flows that alter water quality in downstream estuaries
\citep{bricker_effects_2008, kennicuttWaterQualityGulf2017, bugica_water_2020}.
Despite these increasing pressures, national scale assessments have
classified coastal estuaries in Texas as moderate or lower for
exhibiting eutrophic conditions \citep{bricker_effects_2008}. However, a
suite of recent studies indicate that estuary water quality dynamics in
both agriculturally dominated and urban watersheds within Texas are in
fact expressing conditions that are increasingly conducive to algal
blooms and eutrophication
\citep{wetzWaterQualityDynamics2016, wetz_exceptionally_2017, bugica_water_2020, chinPhytoplanktonBiomassCommunity2022}.
With identification of localized areas of estuary water quality concern
along the Texas coast \citep{bugica_water_2020}, localized studies are
being prioritized to better inform management actions.

This project provides an assessment of nutrient loading and water
quality responses in Lavaca Bay, Texas. Lavaca Bay is a secondary bay in
the larger Matagorda Bay system located roughly halfway between Houston,
Texas and Corpus Christi, Texas. Lavaca Bay has faced substantial
challenges associated with legacy contamination but general water
quality parameters such as dissolved oxygen (DO), nutrients, and
biological parameters have been well within state water quality
standards. However, long-term declines of benthic fauna abundance,
biomass, and diversity in Lavaca Bay that are linked to reductions in
freshwater inflows and changes in estuary salinity
\citep{beserespollackLongtermTrendsResponse2011, palmerImpactsDroughtsLow2015, montagnaAssessmentRelationshipFreshwater2020}
and are a concern to local stakeholders. \citet{bugica_water_2020}
identified monotonic increases in total phosphorus (TP), orthophosphate,
total Kjeldahl nitrogen (TKN), and chlorophyll-\emph{a} at sites within
Lavaca Bay. Although no long-term changes in DO concentrations were
identified, the trends in nutrient concentrations are concerning due to
the role of nitrogen as a limiting factor for primary production in many
Texas estuaries
\citep{gardnerNitrogenFixationDissimilatory2006, houTransformationFateNitrate2012, doradoUnderstandingInteractionsFreshwater2015, paudelRelationshipSuspendedSolids2019, wetz_exceptionally_2017}
and the ramifications that major changes in nitrogen loadings could have
for productivity and eutrophication in Lavaca Bay.

There are ongoing efforts between local, state, and federal agencies to
address water quality impairments in the freshwater portions of the
Lavaca Bay watershed
\citep{jainTechnicalSupportDocument2021, schramm_lavaca_2018, bertholdDirectMailingEducation2021}.
However, at a statewide scale, these approaches have shown limited
success and emphasize a need for improved efforts at assessing and
linking management actions with downstream water quality to better
identify and replicate effective management actions across the state
\citep{schrammTotalMaximumDaily2022}. The identification and
communication of changes and trends in water quality is complicated by
the fact that trends are often non-linear and confounded by
precipitation and runoff that hinder traditional analysis
\citep{wazniakLinkingWaterQuality2007, lloydMethodsDetectingChange2014}.
To provide actionable information for resource managers, water quality
conditions must be evaluated relative to changes in natural
environmental drivers to better understand and manage potential
anthropogenic effects.

Here something some we present methodology and results for estimating
nutrient loading- and assessing linkages in estuary water quality
parameters with watershed derived nutrient loads\ldots.

@@

\begin{figure}
\begin{adjustwidth}{-\extralength}{0cm}
\centering
\includegraphics[width=15.5cm]{Schramm-Manuscript-2023_files/figure-latex/fig1-1.png}
\end{adjustwidth}
\caption{Map of the Lavaca Bay watershed, location of USGS gages where nutrient loads were calculated, and location of estuary water quality sampling sites.\label{fig1}}
\end{figure}

Lavaca Bay is a secondary bay in the Matagorda Bay system located on the
Texas Gulf coast, roughly halfway between the cities of Houston and
Corpus Christi (Figure \ref{fig:fig1}). Lavaca Bay is 190
km\textsuperscript{2} with the majority of freshwater inflow provided by
the Lavaca and Navidad River systems. The Garcitas-Arenosa, Placedo
Creek, and Cox Bay watersheds provide additional freshwater inflows. The
entire watershed land area for Lavaca Bay is 8,149
km\textsuperscript{2}. The Lavaca and Navidad River watersheds are a
combined 5,966 km\textsuperscript{2}, or approximately 73\% of the
entire Lavaca Bay watershed area. Discharge from the Navidad River is
regulated by Lake Texana which has been in operation since 1980. Lake
Texana provides 170,000 acre-feet of water storage and discharges into
the tidal section of the Navidad River which ultimately joins the tidal
section of the Lavaca River 15 km upstream of the confluence with the
Bay.

\textbf{Need to add a short description of land use}

@@

Water quality sample data for the two freshwater and three estuary
locations were obtained from the Texas Commission on Environmental
Quality (TCEQ) Surface Water Quality Monitoring Information System. Data
submitted through the system are required to be collected under Quality
Assurance Project Plans and lab method procedures outlined by the TCEQ's
procedures manual. The QAPP and procedures manuals ensure the consistent
collection and laboratory methods are applied between samples collected
by different entities and under different projects. All sites had
varying lengths of and availability of data. For freshwater locations,
TP from January 2000 through December 2020 and nitrate-nitrogen
(NO\textsubscript{3}) data from January 2005 through December 2020 were
downloaded (Table \ref{tab:fwsummary}). Less than 5-years of total
nitrogen and TKN concentration data were available at the freshwater
sites and deemed insufficient to develop load estimation models
(\textbf{CITE}). The three estuary sites included an upper Lavaca Bay
site near the outlet of the Lavaca River system (TCEQ-13563), a
mid-Lavaca Bay site (TCEQ-13383), and the lower Lavaca Bay site near the
mouth of the Bay (TCEQ-13384). For estuary locations, we obtained data
for TP, Nitrite+Nitrate (NO\emph{\textsubscript{x}}), TKN,
chlorophyll-\emph{a}, and DO concentrations from January 2005 through
December 2020 (Table \ref{tab:estuarysummary}).

@@

Estimates of NO\textsubscript{3} and TP loads at the Lavaca River
(USGS-08164000) and the outlet of Lake Texana on the Navidad River
(USGS-08164250) were developed using Generalized Additive Models (GAMs)
relating nutrient concentration to river discharge, season, and time.
Separate models were fit at each station for each parameter and used to
predict nutrient concentrations for each day in the study period. GAMs
can be specified in a functionally similar manner to the commonly used
LOADEST \citep{cohn_validity_1992} or WRTDS \citep{hirsch_weighted_2010}
regression models and have been shown to produce reliable estimates of
nutrient and sediment loadings
\citep{wangLoadEstimationUncertainties2011, kroonRiverLoadsSuspended2012, kuhnert_quantifying_2012, robson_prediction_2015-1, hagemannEstimatingNutrientOrganic2016, mcdowell_implications_2021, biagi_novel_2022}.
GAMs are a semiparametric extension of generalized linear models where
the linear predictor is represented as the sum of multiple unknown
smooth functions and parametric linear predictors
\citep{wood_fast_2011}. Although the underlying parameter estimation
procedure of GAMs is substantially different than WRTDS, both the
functional form and results are demonstrated to be similar
\citep{beckNumericalQualitativeContrasts2017}. The use of GAMs over
other regression-based approaches was (1) the ability to easily explore
and incorporate different model terms, (2) the ability to incorporate
non-linear smooth function without explicit apriori knowledge of the
expect shape, and (3) the ability to specify a link function that
relates the expected value of the response to the linear predictors and
allows use to avoid data transformations as much as possible.

GAMs were fit using the \emph{mgcv} package in R which makes available
multiple types of smooth functions with automatic smoothness selection
\citep{wood_fast_2011}. The general form of the model relating
NO\textsubscript{3} and TP concentration to streamflow, season, and time
was:

@@

where \(\mu\) is the conditional expected NO\textsubscript{3}-N or TP
concentration, \emph{g()} is the log-link, \(\alpha\) is the intercept,
\emph{f\textsubscript{n}()} are smoothing functions. \emph{y} is the
response variable (NO\textsubscript{3} or TP concentration) modeled as
normally distributed with mean \(\mu\) and standard deviation
\(\sigma\). \emph{ddate} is the date converted to decimal notation,
\emph{yday} is numeric day of year (1-366), and \emph{log1p(Q)} is the
natural log of mean daily streamflow plus 1.

Moving average (\emph{ma}) is an exponentially smoothed moving average
that attempts to incorporate the influence of prior streamflow events on
concentration at the current time period.
\citet{wangLoadEstimationUncertainties2011},
\citet{kuhnert_quantifying_2012} and \citet{zhang_improving_2017} refer
to this as averaged or smoothed discounted flow and demonstrated
improvements in nutrient loading models by including the term.
\citet{kuhnert_quantifying_2012} expresses \emph{ma} as:

@@

where \emph{x} are the averages of log-transformed streamflow over the
antecedent period (\emph{1-year}, \emph{1-month}, etc.) for time
\emph{t}. We used \emph{ltfa} in NO\textasciitilde3 models and
\emph{stfa} in TP models based on results from
\citet{zhang_improving_2017} demonstrating major improvements in
NO\textsubscript{x} regression models that incorporated \emph{ltfa} and
moderate improvements in TP regression models that incorporated
\emph{stfa}.

@@

Daily loads were estimated as the predicted concentration multiplied by
the daily streamflow. For the Lake Texana site, model terms were
slightly modified because daily loads are a function of dam releases and
nutrient concentration, but concentration will be a function of lake
inflows and or other lake processes. \emph{Q}, \emph{ma}, and \emph{fa}
terms were calculated based on total gaged inflow from the 4 major
tributaries to the lake and laily loads at the dam were calculated from
the discrete daily concentration at the discharge point of the lake and
corresponding reported daily discharge from the dam. Flow-normalized
loads were estimated similar to WRTDS by setting flow-based covariates
on each day of the year equal to each of the historical values for that
day of the year over the study period \citep{hirsch_weighted_2010}. The
flow-normalized estimate was calculated as the mean of all the
predictions for each day considering all possible flow values. Standard
deviations and credible intervals were obtained by drawing samples from
the multivariate normal posterior distribution of the fitted GAM
\citep{woodConfidenceIntervalsGeneralized2006, marraCoveragePropertiesConfidence2012, mcdowell_implications_2021}.
Uncertainty in loads were reported as 90\% credible intervals developed
by drawing 1000 realizations of parameter estimates from the
multivariate normal posterior distribution of the model parameters. GAM
performance was evaluated using repeated 5-fold cross validation
\citep{burmanComparativeStudyOrdinary1989} and average Nash-Suttclifee
Efficiency (NSE), r\textsuperscript{2} and percent bias (PBIAS) metrics
across folds were calculated for each model.

@@

where \(\mu\) is the conditional expected response (nutrient
concentration), \emph{g()} is the log link, and response variable was
modeled as Gamma distributed with mean \(\mu\) and scale \(\lambda\).
\emph{f\textsubscript{1}(ddate)} is decimal date smoothed with a
thin-plate regression spline, \emph{f\textsubscript{2}(yday)} is the
numeric day of year smoothed with a cyclic cubic regression spline,
\emph{f\textsubscript{3}(Q)} is mean daily inflow (the combined
measurements from Lavaca River and Lake Texana) and
\emph{f\textsubscript{4}(Load)} is the total NO\textsubscript{3} or TP
watershed load. The set of models specified for each water quality
response are in Table \ref{rab:estgammodels}.

@@

Because streamflow and nutrient loads are tightly correlated, freshwater
inflow can mask signals from nutrient loads alone. Following the
methodology implemented by
\citet{murphyNutrientImprovementsChesapeake2022}, both streamflow and
nutrient loads were prepossessed to account for season and flow. Instead
of using raw freshwater inflow and nutrient loading values, these values
were replaced by seasonally adjusted inflow and flow-adjusted nutrient
loads by fitting a GAM relating season (day of year) to log transformed
daily freshwater inflow values:

@@

Based on criteria provided by \citet{moriasiHydrologicWaterQuality2015},
GAMs ranged from ``satisfactory'' to ``very good'' using median NSE,
r\textsuperscript{2} and PBIAS calculated using repeated 5-fold cross
validation on predicted and measured nutrient loads. NO\textsubscript{3}
GAM models had median NSE values of 0.34 and 0.48 at USGS-08164000 and
USGS-08164390 respectively. Median r\textsuperscript{2} values were 0.70
(USGS-08164000) and 0.87 (USGS-08164525) and PBIAS values were 2.00
(USGS-08164000) and 10.90 (USGS-08164525) for NO\textsubscript{3} loads.
Median NSE values were 0.80 (USGS-08164000) and 0.91 (USGS-08164525) for
TP loads. Median r\textsuperscript{2} values for TP loads were 0.93
(USGS-08164000) and 0.99 (USGS-08164525) and median PBIAS values were
-7.20 (USGS-08164000) and -3.30 (USGS-08164525). Density plots of
metrics show similar distribution of values between sites for the same
parameter, with the exception r\textsuperscript{2} values for
NO\textsubscript{3} loads where USGS-08164000 showed a much larger
variance in values compared to USGS-08164525 (Figure \ref{fig:fig2}).In
addition to higher average NSE and r\textsuperscript{2} values, GAMs had
smaller variance in metric values for TP compared to
NO\textsubscript{3}.

@@

GAMs show that increases in freshwater inflow result in nearly linear
increases in TP and NO\textsubscript{\emph{x}} at all three sites
(Figure \ref{fig:6}). At the upper-bay TCEQ-13563 site, GAMs showed
increases in freshwater inflow initially increased chlorophyll-\emph{a}
and DO concentration but concentrations leveled and potentially
decreased at higher flows. The mid-bay TCEQ-13383 site showed a nearly
linear increased in chlorophyll-\emph{a} concentration with increases in
freshwater inflow. Freshwater flow did not have significant effects on
chlorophyll-\emph{a}, TKN, or DO at the lower-bay TCEQ-13384 site.

Increases in TP loads resulted in nearly linear increases of TP
concentration at the upper- and mid-bay sites, TCEQ-13563 and TCEQ-13383
respectively (Figure \ref{fig:fig7}). The relative effect size appeared
to much smaller than the effect of freshwater inflow alone. Increases in
NO\textsubscript{3} loads only showed an effect at the lower-bay
TCEQ-13384 site. The effect was quite small compared to streamflow and
provided only small improvements to the model (Table \ref{tab:tab4}). As
noted above, nutrient loadings did not provide any explanation in
changes in the remaining assessed water quality parameters.

@@

The non-linear temporal water quality trends identified using GAMs
differed slightly from the trends identified by
\citet{bugicaWaterQualityTrends2020}. This is not unexpected due to the
different time periods, different methodology, and generally small
slopes identified for most of the significant water quality parameters
in \citet{bugica_water_2020}. The trend in DO and cholorophyll-\emph{a}
concentrations are stable in comparison to other Texas estuaries that
might be facing larger demands for freshwater diversions, higher
population growth, and more intense agricultural production
\citep{wetzWaterQualityDynamics2016, bugica_water_2020}. The trend of
increasing NO\textsubscript{\emph{x}} concentration at the upper-bay
TCEQ-13563 site and recent increases in TKN concentration at the mid-bay
TCEQ-13383 site are concerning due to the nitrogen limitation identified
in many Texas estuaries
\citep{gardnerNitrogenFixationDissimilatory2006, houTransformationFateNitrate2012, doradoUnderstandingInteractionsFreshwater2015, paudelRelationshipSuspendedSolids2019, wetz_exceptionally_2017}
and the relatively low ambient concentrations observed in Lavaca Bay.

@@

GAMs did not identify responses in DO concentration to inflows or
nutrient loads. The seasonality term in the temporal GAM models
explained a substantial amount of DO variation at all of the sites.
Responses of estuary metabolic processes and resulting DO concentrations
can be quite complicated and often locally specific
\citep{caffreyFactorsControllingNet2004}. While the lack of total
nitrogen or TKN loading data hinders interpretation, the large seasonal
effect on DO suggests physical factors play an important role and should
be included in future models. Prior work suggests that there is some
evidence the Lavaca Bay may not be limited by nutrients alone, with high
turbidity or nutrient processing in upper portions of the Bay or
intertidal river limiting production \citep{russell_effect_2006}.

@@

This section is not mandatory, but can be added to the manuscript if the
discussion is unusually long or complex.

@@

\funding{This paper was funded by financial assistance provided by the
Coastal Zone Management Act of 1972, as amended, administered by the
National Oceanic and Atmospheric Administration (NOAA), Office for
Coastal Management, pursuant to NOAA Award No.~NA21NOS4190136. The views
expressed herein are those of the author(s) and do not necessarily
reflect the views of NOAA, the U.S. Department of Commerce, or any of
their subagencies.}

>>>>

% Abstract (Do not insert blank lines, i.e. \\)
\abstract{Lavaca Bay is a small secondary embayment on the Texas coast
that is displaying early signals of water quality degradation. This
study uses Generalized Additive Models to assesses both watershed
nutrient loading and responses in estuary water quality to nutrient
loading and streamflow. While delivered nutrient loads varied greatly,
flow-normalized nutrient loads indicated limited evidence of changes in
nutrient runoff in the watershed since 2005. As a consequence, there
wasn't strong evidence for an effect of nutrient loading (controlled for
flow) on dissolved oxygen or chlorophyll-\emph{a} concentrations. There
was evidence total phosphorus (TP) loading had a strong positive effect
on TP concentration in the Bay, particularly at sites closest to the
river mouth. Seasonally-adjusted streamflow had strong effects on
nutrient concentration and chlorophyll-\emph{a} concentrations at most
sites. Although temporal trends did not identify significant estuary
water quality degredation in dissolved oxygen or chlorophyll-\emph{a}
concentrations, site specific trends in both organic and inorganic
nitrogen are concerning for the long-term health of the estuary. This
study provides a baseline assessment of the linkages between watershed
nutrient loading and water quality responses in Lavaca Bay. Some of the
data gaps highlighted in the study are being addressed in ongoing
projects to provide improved information for resource managers.}

@@

Like many coastal areas globally, coastal watersheds along the Texas
Gulf coast are facing pressures from increasing population, increases in
point source and non-point source pollution and alterations to
freshwater flows that affect water quality in downstream estuaries
\citep{bricker_effects_2008, kennicuttWaterQualityGulf2017, bugica_water_2020}.
Despite these increasing pressures, national scale assessments have
classified coastal estuaries in Texas as moderate or lower for
exhibiting eutrophic conditions \citep{bricker_effects_2008}. However, a
suite of recent studies indicate that estuary water quality dynamics in
both agriculturally dominated and urban watersheds within Texas are in
fact expressing conditions that are increasingly conducive to algal
blooms and eutrophication
\citep{wetzWaterQualityDynamics2016, wetz_exceptionally_2017, bugica_water_2020, chinPhytoplanktonBiomassCommunity2022}.
With identification of localized areas of estuary water quality concern
along the Texas coast \citep{bugica_water_2020}, localized studies are
being prioritized to better inform management actions.

This project provides an assessment of nutrient loading and water
quality responses in Lavaca Bay, Texas. Lavaca Bay is a secondary bay in
the larger Matagorda Bay system located roughly halfway between Houston,
Texas and Corpus Christi, Texas. Lavaca Bay has faced substantial
challenges associated with legacy contamination but general water
quality parameters such as dissolved oxygen (DO), nutrients, and
biological parameters have been well within state water quality
standards. However, long-term declines of benthic fauna abundance,
biomass, and diversity in Lavaca Bay primarily linked to reductions in
freshwater inflows and changes in estuary salinity
\citep{beserespollackLongtermTrendsResponse2011, palmerImpactsDroughtsLow2015, montagnaAssessmentRelationshipFreshwater2020}
are a concern to local stakeholders. \citet{bugica_water_2020}
identified monotonic increases in total phosphorus (TP), orthophosphate,
total Kjeldahl nitrogen (TKN), and chlorophyll-\emph{a} at sites within
Lavaca Bay. Although long-term changes in DO concentrations were not
identified, the trends in nutrient concentrations are concerning due to
the role of nitrogen as a limiting factor for primary production in many
Texas estuaries
\citep{gardnerNitrogenFixationDissimilatory2006, houTransformationFateNitrate2012, doradoUnderstandingInteractionsFreshwater2015, paudelRelationshipSuspendedSolids2019, wetz_exceptionally_2017}
and the ramifications that changes in nitrogen loadings could have for
productivity and eutrophication in Lavaca Bay.

There are ongoing efforts between local, state, and federal agencies to
address water quality impairments in the freshwater portions of the
Lavaca Bay watershed
\citep{jainTechnicalSupportDocument2021, schramm_lavaca_2018, bertholdDirectMailingEducation2021}.
However, at a statewide scale, these approaches have shown limited
success and emphasize a need for improved efforts at assessing and
linking management actions with downstream water quality to better
identify and replicate effective management actions across the state
\citep{schrammTotalMaximumDaily2022}. The identification and
communication of changes and trends in water quality is complicated by
the fact that trends are often non-linear and confounded by
precipitation and runoff that hinder traditional analysis
\citep{wazniakLinkingWaterQuality2007, lloydMethodsDetectingChange2014}.
To provide actionable information for resource managers, water quality
conditions must be evaluated relative to changes in natural
environmental drivers to better understand and manage potential
anthropogenic effects. This study utilizes semiparemetric methods to
develop estimates of delivered and flow-normalized nutrient loads and
assess changes in loads delivered to Lavaca Bay. The study also assesses
the response of water quality parameters in Lavaca Bay over time and in
response to freshwater inflow controlled for seasonality and to
watershed nutrient loads that are controlled for environmentally driven
variation.

@@

\begin{figure}
\includegraphics[width=1\linewidth]{Schramm-Manuscript-2023_files/figure-latex/fig1-1.png}
\caption{Map of the Lavaca Bay watershed, location of USGS gages where nutrient loads were calculated, and location of estuary water quality sampling sites.\label{fig:fig1}}
\end{figure}

Lavaca Bay is 190 km\textsuperscript{2} with the majority of freshwater
inflow provided by the Lavaca and Navidad River systems (Figure
\ref{fig:fig1}). The Garcitas-Arenosa, Placedo Creek, and Cox Bay
watersheds provide additional freshwater inflows. The entire watershed
land area is 8,149 km\textsuperscript{2} and primarily rural. Watershed
land cover and land use is 50\% grazed pasture and rangeland, 20\%
cultivated cropland (primarily rows crops such as corn, cotton, and
sorghum), and 5\% suburban/urban. Pasture and rangeland is concentrated
in the Lavaca River watershed, while cultivated crops are generally
located along the eastern tributaries of the Navidad river. The Lavaca
and Navidad River watersheds are a combined 5,966 km\textsuperscript{2},
or approximately 73\% of the entire Lavaca Bay watershed area. Discharge
from the Navidad River is regulated by Lake Texana which has been in
operation since 1980. Lake Texana provides 0.210 km\textsuperscript{3}
of water storage and discharges into the tidal section of the Navidad
River which ultimately joins the tidal section of the Lavaca River 15 km
upstream of the confluence with the Lavaca Bay.

@@

Water quality sample data for the two freshwater and three estuary
locations were obtained from the Texas Commission on Environmental
Quality (TCEQ) Surface Water Quality Monitoring Information System. Data
submitted through the system are required to be collected under Quality
Assurance Project Plans and lab method procedures outlined by the TCEQ's
procedures manual. The QAPP and procedures manuals ensure the consistent
collection and laboratory methods are applied between samples collected
by different entities and under different projects. All sites had
varying lengths of and availability of data. For freshwater locations,
TP from January 2000 through December 2020 and nitrate-nitrogen
(NO\textsubscript{3}) data from January 2005 through December 2020 were
downloaded (Table \ref{tab:fwsummary}). Less than 5-years of total
nitrogen and TKN concentration data were available at the freshwater
sites and deemed insufficient to develop load estimation models
\citep{horowitzEvaluationSedimentRating2003, snelderEstimationCatchmentNutrient2017}.
The three estuary sites included an upper Lavaca Bay site near the
outlet of the Lavaca River system (TCEQ-13563), a mid-Lavaca Bay site
(TCEQ-13383), and the lower Lavaca Bay site near the mouth of the Bay
(TCEQ-13384). For estuary locations, we obtained data for TP,
Nitrite+Nitrate (NO\emph{\textsubscript{x}}), TKN, chlorophyll-\emph{a},
and DO concentrations from January 2005 through December 2020 (Table
\ref{tab:estuarysummary}).

@@

Estimates of NO\textsubscript{3} and TP loads at the Lavaca River
(USGS-08164000) and the outlet of Lake Texana on the Navidad River
(USGS-08164525) were developed using Generalized Additive Models (GAMs)
relating nutrient concentration to river discharge, season, and time.
Separate models were fit at each station for each parameter and used to
predict nutrient concentrations for each day in the study period. GAMs
can be specified in a functionally similar manner to the commonly used
LOADEST \citep{cohn_validity_1992} or WRTDS \citep{hirsch_weighted_2010}
regression models and have been shown to produce reliable estimates of
nutrient and sediment loadings
\citep{wangLoadEstimationUncertainties2011, kroonRiverLoadsSuspended2012, kuhnert_quantifying_2012, robson_prediction_2015-1, hagemannEstimatingNutrientOrganic2016, mcdowell_implications_2021, biagi_novel_2022}.
GAMs are a semiparametric extension of generalized linear models where
the linear predictor is represented as the sum of multiple unknown
smooth functions and parametric linear predictors
\citep{wood_fast_2011}. Although the underlying parameter estimation
procedure of GAMs is substantially different than WRTDS, both the
functional form and results are demonstrated to be similar
\citep{beckNumericalQualitativeContrasts2017}. The use of GAMs over
other regression-based approaches was (1) the ability to easily explore
and incorporate different model terms, (2) the ability to incorporate
non-linear smooth function without explicit a priori knowledge of the
expect shape, and (3) the ability to specify a link function that
relates the expected value of the response to the linear predictors and
avoid data transformations.

GAMs were fit using the \emph{mgcv} package in R which makes available
multiple types of smooth functions with automatic smoothness selection
\citep{wood_fast_2011}. The general form of the model related
NO\textsubscript{3} or TP concentration to a long term tend, season,
streamflow, and two different antecedent discharge terms:

@@

where \(\mu\) is the conditional expected NO\textsubscript{3} or TP
concentration, \emph{g()} is the log-link, \(\alpha\) is the intercept,
\emph{f\textsubscript{n}()} are smoothing functions. \emph{y} is the
response variable (NO\textsubscript{3} or TP concentration) modeled as
normally distributed with mean \(\mu\) and standard deviation
\(\sigma\). \emph{ddate} is the date converted to decimal notation,
\emph{yday} is numeric day of year (1-366), and \emph{log1p(Q)} is the
natural log of mean daily streamflow plus 1.

Moving average (\emph{ma}) is an exponentially smoothed moving average
that attempts to incorporate the influence of prior streamflow events on
concentration at the current time period
\citep{wangLoadEstimationUncertainties2011, kuhnert_quantifying_2012, zhang_improving_2017},
and expressed in \citet{kuhnert_quantifying_2012} as:

@@

where \emph{x} are the averages of log-transformed streamflow over the
antecedent period (\emph{1-year}, \emph{1-month}, etc.) for time
\emph{t}. We used \emph{ltfa} in NO\textsubscript{3} models and
\emph{stfa} in TP models based on results from
\citet{zhang_improving_2017} demonstrating major improvements in
NO\textsubscript{x} regression models that incorporated \emph{ltfa} and
moderate improvements in TP regression models that incorporated
\emph{stfa}.

@@

Daily loads were estimated as the predicted concentration multiplied by
the daily streamflow. For the Navidad River (USGS-08164525) site, daily
loads at the dam were calculated from the discrete daily concentration
at the discharge point of the lake and corresponding reported daily
discharge from the dam. Flow-normalized loads were estimated similar to
WRTDS by setting flow-based covariates on each day of the year equal to
each of the historical values for that day of the year over the study
period \citep{hirsch_weighted_2010}. The flow-normalized estimate was
calculated as the mean of all the predictions for each day considering
all possible flow values. Standard deviations and credible intervals
were obtained by drawing samples from the multivariate normal posterior
distribution of the fitted GAM
\citep{woodConfidenceIntervalsGeneralized2006, marraCoveragePropertiesConfidence2012, mcdowell_implications_2021}.
Uncertainty in loads were reported as 90\% credible intervals developed
by drawing 1000 realizations of parameter estimates from the
multivariate normal posterior distribution of the model parameters. GAM
performance was evaluated using repeated 5-fold cross validation
\citep{burmanComparativeStudyOrdinary1989} and average Nash-Sutcliffe
Efficiency (NSE), r\textsuperscript{2} and percent bias (PBIAS) metrics
across folds were calculated for each model.

@@

where \(\mu\) is the conditional expected response (nutrient
concentration), \emph{g()} is the log link, and response variable was
modeled as Gamma distributed with mean \(\mu\) and scale \(\lambda\).
\emph{f\textsubscript{1}(ddate)} is decimal date smoothed with a
thin-plate regression spline, \emph{f\textsubscript{2}(yday)} is the
numeric day of year smoothed with a cyclic cubic regression spline,
\emph{f\textsubscript{3}(Q)} is mean daily inflow (the combined
measurements from Lavaca River and Navidad River) and
\emph{f\textsubscript{4}(Load)} is the total NO\textsubscript{3} or TP
watershed load. The set of models specified for each water quality
response are in Table \ref{tab:estgammodels}.

@@

Because streamflow and nutrient loads are tightly correlated, freshwater
inflow can mask signals from nutrient loads alone. Following the
methodology implemented by
\citet{murphyNutrientImprovementsChesapeake2022}, both streamflow and
nutrient loads were prepossessed to account for season and flow.
Freshwater inflow and nutrient loads were replaced by seasonally
adjusted log transformed inflow and flow-adjusted log transformed
nutrient loads obtained by fitting a GAM relating season (day of year)
to log transformed daily freshwater inflow values:

@@

Based on criteria provided by \citet{moriasiHydrologicWaterQuality2015},
GAMs ranged from ``satisfactory'' to ``very good'' using median NSE,
r\textsuperscript{2} and PBIAS calculated using repeated 5-fold cross
validation on predicted and measured nutrient loads. NO\textsubscript{3}
GAM models had median NSE values of 0.34 and 0.48 at USGS-08164000 and
USGS-08164390 respectively. Median r\textsuperscript{2} values were 0.70
(USGS-08164000) and 0.87 (USGS-08164525) and PBIAS values were 2.00
(USGS-08164000) and 10.90 (USGS-08164525) for NO\textsubscript{3} loads.
Median NSE values were 0.80 (USGS-08164000) and 0.91 (USGS-08164525) for
TP loads. Median r\textsuperscript{2} values for TP loads were 0.93
(USGS-08164000) and 0.99 (USGS-08164525) and median PBIAS values were
-7.20 (USGS-08164000) and -3.30 (USGS-08164525). Density plots of
metrics show similar distribution of values between sites for the same
parameter, with the exception r\textsuperscript{2} values for
NO\textsubscript{3} loads where USGS-08164000 showed a much larger
variance in values compared to USGS-08164525 (Figure \ref{fig:fig2}). TP
GAMS had higher average NSE and r\textsuperscript{2} values and lower
variance in metric values compared to NO\textsubscript{3}.

@@

GAMs showed increases in freshwater inflow resulted in nearly linear
increases in TP and NO\textsubscript{\emph{x}} concentration at all
three sites (Figure \ref{fig:fig6}). At the upper-bay TCEQ-13563 site,
GAMs showed increases in freshwater inflow initially increased
chlorophyll-\emph{a} and DO concentration but concentrations leveled and
potentially decreased at higher flows. The mid-bay TCEQ-13383 site
showed a nearly linear increased in chlorophyll-\emph{a} concentration
in response to increases freshwater inflow. Freshwater flow did not have
significant effects on chlorophyll-\emph{a}, TKN, or DO at the lower-bay
TCEQ-13384 site.

Increased TP loads resulted in nearly linear increases of TP
concentration at the upper- and mid-bay sites, TCEQ-13563 and TCEQ-13383
respectively (Figure \ref{fig:fig7}). The relative effect size appeared
to much smaller than the effect of freshwater inflow alone. Increased
NO\textsubscript{3} loads only showed an effect at the lower-bay
TCEQ-13384 site. The effect was quite small compared to streamflow and
provided only small improvements to the model (Table \ref{tab:tab4}). As
noted above, nutrient loadings did not provide any explanation in
changes in the remaining assessed water quality parameters.

@@

The non-linear temporal water quality trends identified using GAMs
differed slightly from the trends identified by
\citet{bugica_water_2020}. This is not unexpected due to the different
time periods, different methodology, and generally small slopes
identified for most of the significant water quality parameters in prior
work. The trend in DO and cholorophyll-\emph{a} concentrations are
stable in comparison to other Texas estuaries that might be facing
larger demands for freshwater diversions, higher population growth, and
more intense agricultural production
\citep{wetzWaterQualityDynamics2016, bugica_water_2020}. The trend of
increasing NO\textsubscript{\emph{x}} concentration at the upper-bay
TCEQ-13563 site and recent increases in TKN concentration at the mid-bay
TCEQ-13383 site are concerning due to the nitrogen limitation identified
in many Texas estuaries
\citep{gardnerNitrogenFixationDissimilatory2006, houTransformationFateNitrate2012, doradoUnderstandingInteractionsFreshwater2015, paudelRelationshipSuspendedSolids2019, wetz_exceptionally_2017}
and the relatively low ambient concentrations observed in Lavaca Bay.

@@

GAMs did not identify responses in DO concentration to inflows or
nutrient loads. The seasonality term in the temporal GAM models
explained a substantial amount of DO variation at all of the sites.
Responses of estuary metabolic processes and resulting DO concentrations
can be quite complicated and often locally specific
\citep{caffreyFactorsControllingNet2004}. While the lack of total
nitrogen or TKN loading data hinders interpretation, the large seasonal
effect on DO suggests physical factors play an important role and should
be included in future models. Prior work suggests that Lavaca Bay may
not be limited by nutrients alone, with high turbidity or nutrient
processing in upper portions of the Bay or intertidal river limiting
production \citep{russell_effect_2006}.

@@

GAM models appear to provide reliable estimates of nutrient loads in the
Lavaca Bay watershed. However, additional flow-biased data collection
efforts would decrease the prediction variance and probably improve
accuracy at critical high flow events. Ongoing projects will fill data
gaps for total nitrogen and TKN loading. This study, consistent with
others along the Texas coast, found strong effects of freshwater flow on
nutrient and chlorophyll-\emph{a} concentrations. DO concentrations,
dominated by seasonal effects, did not show strong direct responses to
freshwater flow. Small variance in flow-adjusted nutrient loads
indicates that (1) there have been limited changes in non point sources
of nutrients and (2) that those changes have limited effect on water
quality in Lavaca Bay to date. Although the study did not identify
strong responses to changes in nutrient loading, this does provide a
baseline assessment for future water quality management activities in
the watershed.

@@

\funding{This project was funded in part by a Texas Coastal Management
Program grant approved by the Texas Land Commissioner, providing
financial assistance under the Coastal Zone Management Act of 1972, as
amended, awarded by the National Oceanic and Atmospheric Administration
(NOAA), Office for Coastal Management, pursuant to NOAA Award
No.~NA21NOS4190136. The views expressed herein are those of the
author(s) and do not necessarily reflect the views of NOAA, the U.S.
Department of Commerce, or any of their subagencies}
